---
title: "Time Series Analysis"
author: "Katarzyna Gworek"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: hide
---

# Project overview

This report analyses two monthly time series from January 2012 to December 2022. The first is the USD/PLN exchange rate, while the second describes greenhouse gas consumption in Italy. The analysis combines data visualization, decomposition, exponential smoothing and ARIMA/SARIMA forecasting.

This report presents the analysis of two monthly time series covering January 2012–December 2022.

The first series describes the USD/PLN exchange rate. The second describes greenhouse gas consumption in Italy. Both series were decomposed in JDemetra+ and subsequently analysed using extrapolation models and ARIMA/SARIMA models.

# 1. Non-seasonal time series – USD/PLN

## 1.1 Data preparation

The USD/PLN data are converted into a monthly time series. The last three observations are kept as an out-of-sample period for forecast evaluation.

```{r setup-nonseasonal, message=FALSE, warning=FALSE}
library(foreign)
library(xts)
library(urca)
library(lmtest)
library(fBasics)
library(forecast)
library(tseries)
library(fUnitRoots)

ASC_niesezonowe <- read.csv(
  "dane/Gworek_ASC_niesezonowe.csv",
  header = TRUE, sep = ";", dec = ","
)

ASC_niesezonowe.ts <- ts(
  ASC_niesezonowe$rate, start = c(2012, 1), frequency = 12
)

ASC_niesezonowe.in <- window(ASC_niesezonowe.ts, end = c(2022, 9))
ASC_niesezonowe.out <- window(ASC_niesezonowe.ts, start = c(2022, 10))
```

The training sample contains observations up to September 2022, while October–December 2022 are used to test forecast accuracy.

## 1.2 Visualization

The plot is used to assess the general behaviour of the exchange rate and to check for visible trend or seasonal patterns. The series does not show a clear seasonal pattern.

```{r plot-nonseasonal}
plot(ASC_niesezonowe.ts,
     main = "USD/PLN exchange rate",
     xlab = "Date", ylab = "Exchange rate")
```

## 1.3 Exponential smoothing – EWMA

EWMA is used as a simple benchmark model without trend or seasonality. It provides a short-term forecast based mainly on the recent level of the series.

```{r ewma}
EWMA <- HoltWinters(ASC_niesezonowe.in, beta = FALSE, gamma = FALSE)

EWMA.forecast <- predict(
  EWMA, n.ahead = 3, prediction.interval = TRUE
)
EWMA.forecast
```

```{r ewma-plot}
plot(ASC_niesezonowe.ts, main = "EWMA forecast",
     xlab = "Date", ylab = "Exchange rate")
lines(EWMA.forecast[, 1])
lines(EWMA.forecast[, 2], lty = 2)
lines(EWMA.forecast[, 3], lty = 2)
abline(v = 2022 + 9/12, lty = 2)
```

## 1.4 Holt model

The Holt model extends exponential smoothing by allowing for a trend. It is therefore compared with EWMA because the exchange rate shows a general upward movement over the analysed period.

```{r holt}
Holt <- HoltWinters(ASC_niesezonowe.in, gamma = FALSE)

Holt.forecast <- predict(
  Holt, n.ahead = 3, prediction.interval = TRUE
)
Holt.forecast
```

```{r holt-plot}
plot(ASC_niesezonowe.ts, main = "Holt forecast",
     xlab = "Date", ylab = "Exchange rate")
lines(Holt.forecast[, 1])
lines(Holt.forecast[, 2], lty = 2)
lines(Holt.forecast[, 3], lty = 2)
abline(v = 2022 + 9/12, lty = 2)
```

## 1.5 ARIMA analysis

The first differences are examined to obtain a stationary series. ACF and PACF are then used to identify plausible AR and MA orders before comparing alternative ARIMA specifications.

```{r arima-preparation}
diff.ASC_niesezonowe.in <- diff.xts(ASC_niesezonowe.in)

plot(diff.ASC_niesezonowe.in, type = "l",
     main = "First differences of USD/PLN")

Acf(diff.ASC_niesezonowe.in, lag.max = 24)
Pacf(diff.ASC_niesezonowe.in, lag.max = 24)
```

```{r arima-models}
arima313 <- arima(diff.ASC_niesezonowe.in, order = c(3, 0, 3))
arima213 <- arima(diff.ASC_niesezonowe.in, order = c(2, 0, 3))
arima113 <- arima(diff.ASC_niesezonowe.in, order = c(1, 0, 3))
arima112 <- arima(diff.ASC_niesezonowe.in, order = c(1, 0, 2))

AIC(arima313); BIC(arima313)
AIC(arima213); BIC(arima213)
AIC(arima112); BIC(arima112)
```

## 1.6 Residual diagnostics

Residual ACF, PACF and Ljung–Box/Box–Pierce tests are used to check whether the selected ARIMA model has adequately captured the dependence in the data.

```{r arima-diagnostics}
Acf(resid(arima112), lag.max = 24)
Pacf(resid(arima112), lag.max = 24)

Box.test(resid(arima112), type = "Ljung-Box", lag = 24)
Box.test(resid(arima112), type = "Box-Pierce", lag = 24)
```

# 2. Seasonal time series – greenhouse gas consumption in Italy

The second series has a clear seasonal pattern: consumption is generally higher in winter and lower during spring and summer. This makes seasonal forecasting methods appropriate.

## 2.1 Data preparation

The Italian greenhouse gas consumption data are converted into a monthly time series. Twelve observations from 2022 are reserved as the out-of-sample period, representing one complete seasonal cycle.

```{r setup-seasonal, message=FALSE, warning=FALSE}
ASC_sezonowe <- read.csv(
  "dane/Gworek_ASC_sezonowe.csv",
  header = TRUE, sep = ";"
)

ASC_sezonowe.ts <- ts(
  ASC_sezonowe$Value, start = c(2012, 1), frequency = 12
)

sezonowy.in <- window(ASC_sezonowe.ts, end = c(2021, 12))
sezonowy.out <- window(ASC_sezonowe.ts, start = c(2022, 1))
```

A full year is reserved for testing so that the forecast evaluation covers all months of the seasonal cycle.

## 2.2 Visualization

The time-series plot is used to identify the main trend and recurring seasonal fluctuations. The visible annual pattern supports the use of seasonal models.

```{r seasonal-plot}
plot(ASC_sezonowe.ts,
     main = "Greenhouse gas consumption in Italy",
     xlab = "Date", ylab = "TJ (GCV)")
```

## 2.3 Holt-Winters models

Both additive and multiplicative Holt-Winters models are estimated because the appropriate form of seasonality is not obvious from the plot alone. Their forecasts can then be compared using forecast errors.

```{r holt-winters}
sezonowy.HWadd <- HoltWinters(sezonowy.in, seasonal = "additive")
sezonowy.HWmult <- HoltWinters(sezonowy.in, seasonal = "multiplicative")

sezonowy.HWadd.forecast <- predict(
  sezonowy.HWadd, n.ahead = 12, prediction.interval = TRUE
)
sezonowy.HWmult.forecast <- predict(
  sezonowy.HWmult, n.ahead = 12, prediction.interval = TRUE
)

sezonowy.HWadd.forecast
sezonowy.HWmult.forecast
```

## 2.4 Logarithmic transformation and differencing

The logarithm is used to reduce the scale of the series. Regular and seasonal differences are examined to obtain a stationary series suitable for SARIMA modelling.

```{r seasonal-differencing}
ln.sezonowy <- log(ASC_sezonowe.ts)
ln.sezonowy.in <- window(ln.sezonowy, end = c(2021, 12))

ln.diff.sezonowy.in <- diff.xts(ln.sezonowy.in, lag = 1)
sdiff.ln.diff.sezonowy.in <- diff.xts(ln.diff.sezonowy.in, lag = 12)

Acf(sdiff.ln.diff.sezonowy.in, lag.max = 36)
Pacf(sdiff.ln.diff.sezonowy.in, lag.max = 36)
```

## 2.5 SARIMA models

SARIMA models account for both non-seasonal and seasonal dependence. Alternative specifications are compared using information criteria and model diagnostics.

```{r sarima}
nobs <- length(ln.sezonowy.in)

arima010111 <- arima(
  ln.sezonowy.in,
  order = c(0, 1, 0),
  seasonal = list(order = c(1, 1, 1), period = 12),
  xreg = 1:nobs
)

arima010011 <- arima(
  ln.sezonowy.in,
  order = c(0, 1, 0),
  seasonal = list(order = c(0, 1, 1), period = 12),
  xreg = 1:nobs
)

AIC(arima010111); BIC(arima010111)
AIC(arima010011); BIC(arima010011)
```

## 2.6 Residual diagnostics

Residual ACF, PACF and Ljung–Box/Box–Pierce tests are used to assess whether the selected SARIMA model adequately captures the remaining autocorrelation.

```{r sarima-diagnostics}
Acf(resid(arima010111), lag.max = 36)
Pacf(resid(arima010111), lag.max = 36)

Box.test(resid(arima010111), type = "Ljung-Box", lag = 36)
Box.test(resid(arima010111), type = "Box-Pierce", lag = 36)
```

# 3. Forecast evaluation

Forecasts are evaluated on the out-of-sample observations using MAE, MSE, MAPE and AMAPE. Lower values indicate more accurate forecasts. For the USD/PLN series, ARIMA(1,1,2) performed best in the original analysis. For Italian greenhouse gas consumption, SARIMA(1,1,1)(0,1,1) achieved the lowest forecast errors among the compared models.

Forecast accuracy is evaluated using MAE, MSE, MAPE and AMAPE. Lower values indicate better forecast performance.

# 4. Conclusions

The results show that the appropriate forecasting method depends on the structure of the series. The non-seasonal USD/PLN series is better described by ARIMA(1,1,2), while the seasonal Italian consumption series is better forecast with SARIMA(1,1,1)(0,1,1).

The analysis combines decomposition, exponential smoothing and ARIMA/SARIMA modelling for two different types of monthly time series: the USD/PLN exchange rate and greenhouse gas consumption in Italy.
